---
title: "New Spatial Bayesian Models"
author: "Raven McKnight"
date: "5/2/2019"
draft: true
ouput: 
  blogdown::html_page:
    toc: true
---

```{r include = FALSE}
library(ggplot2)
library(urbnmapr)
library(wesanderson)
library(tidyverse)
library(CARBayes)
library(sp)
library(ape)
library(spdep)
library(rgdal)
library(spdep)

bc <- read_csv("/Users/raven/Documents/Classes/BayesCapstone/ravenData_Apr17.csv")

counties <- urbnmapr::counties %>%
  mutate(fips = as.numeric(county_fips))  # create field to join on

counties2 <-  st_read('/Users/raven/Documents/tl_2016_us_county', 'tl_2016_us_county')

bc <- bc %>%
  mutate(geoid = substr(GEO.id, start=nchar(GEO.id)-4, stop=nchar(GEO.id))) %>%
  mutate(GEOID = as.factor(geoid))


tn <- bc %>%
  mutate(GEOID=as.factor(fips)) %>%
  filter(State=="Tennessee")

tn.spatial <- merge(x=counties2, y=tn, by="GEOID", all.x=FALSE)

bc.spatial <- merge(counties2, bc, by="GEOID", all.x=FALSE)
bc.sp <- as(bc.spatial, "Spatial")

tn.sp <- sf:::as_Spatial(tn.spatial$geom) 
```


Last week, I fit an initial CARBayes model which identifies boundaries between counties in Tennessee. This week, I'll explore some new models and decide which one(s) to present (and whether I want to fit the model to the entire country!). 

## Back up a second

I've never actually tested to see if there is spatial autocorrelation in my data! I think we can *see* that there is, but we can check with Moran's I to make sure we need to use spatial models. 

```{r}
W.nb <- poly2nb(tn.spatial, row.names=rownames(tn.spatial))   # Create neighborhood matrix
W.list <- nb2listw(W.nb, style="B")
W <- nb2mat(W.nb, style="B")   # We'll need this W for our models below, as well!

moran.mc(tn.spatial$WomenPerClinic, W.list, nsim = 9999) # mc simulations
```

This is kind of surprising! My Moran's I test tells me that the Tennessee data *might* be spatially random. With `observed rank = 8569`, we know that out of 10000 samples, my data is "less random" than about 8500 of them. I'll try Moran's I on the entire country to see if Tennessee is a strange case or if my data really is random. 


```{r cache=TRUE}
st_rook = function(a, b=a) {st_relate(a, b, pattern = "F***1****")}

bc.sp.clean <- bc.spatial %>%    # takes a long time, try not to run again
  mutate(WomenPerClinicClean = replace_na(WomenPerClinic, 0)) %>%
  mutate(NB_ROOK = st_rook(.))

for(i in 1:nrow(bc.sp.clean)){
  bc.sp.clean$nb_length[i] <- length(bc.sp.clean$NB_ROOK[[i]])
}

bc.sp.clean <- bc.sp.clean %>%
  dplyr::select(-NB_ROOK) %>%
  filter(nb_length > 0) 

us.nb <- poly2nb(bc.sp.clean)

us.list <- nb2listw(us.nb, style="B")   

moran.mc(bc.sp.clean$WomenPerClinicClean, us.list, nsim=9999)
```

This makes more sense! Over the entire country, we see that our real data is "less random" than any randomly generated data. This means that we're on the right track to be fitting spatial models. 

## New Models

First, I'll fit a model often used in ecological regression. The goal of this model is to identify factors that affect our response variable--women per clinic. This model is fit using the `S.CARleroux` function in the `CARBayes` packages. Like the model from last week, W is a neighborhood matrix. For this model, I'll include all of my potential covariates to determine which are impactful. 

```{r eval=FALSE}
form <- WomenPerClinic ~ MedianInc + PercRural + PercHSGrad + PercPoverty + PercWhite

new.spatial.model <- S.CARleroux(formula=form, data=tn.spatial, family="gaussian", W=W, burnin=100000, n.sample=300000, thin=20)
```

```{r include = FALSE}
form <- WomenPerClinic ~ MedianInc + PercRural + PercHSGrad + PercPoverty + PercWhite

new.spatial.model <- S.CARleroux(formula=form, data=tn.spatial, family="gaussian", W=W, burnin=100000, n.sample=300000, thin=20)
```
```{r}
print(new.spatial.model)
```

These coefficients are a bit strange because Median Income is on a very different scale than the other predictors. Briefly, here's what we can take away from this model:

* An increase in median income causes an increase in women per clinic -- this seemed counterintuitive to me at first, but we do see that wealthier areas tend to be more populous and therefore more likely to have too few clinics. 
* An increase in percent rural causes a *large* decrease in women per clinic -- this is where the women per clinic metric starts to run into problems. Lots of those rural counties have 0 or 1 clinic, but can't be counted as a desert by our metric because there aren't enough women in need!
* An increase in the proportion of adults with high school diplomas causes an increase in women per clinic. I think this follows the same logic as median income. 
* An increase in the proportion of residents living under the poverty line causes a decrease in women per clinic. Same logic as high school graduates!
* An increase in percent white causes a decrease in women per clinic -- as we might expect, white communities are better served than non-white ones. 

Out of curiosity, I'll fit a model using each race (as collected by the census) as predictors to get a sense of which communities are most underserved. We should keep in mind that this model is fit only on Tennesee which isn't the most diverse state -- it might be better to fit this model on the entire country. 

```{r include = FALSE}
race.form <- WomenPerClinic ~ PercWhite + PercBlack + PercAmInd + PercAsian + PercHawaii + PercTwoOrMore + PercHisp

race.spatial.model <- S.CARleroux(formula=race.form, data=tn.spatial, family="gaussian", W=W, burnin=100000, n.sample=300000, thin=20)
```
```{r eval = FALSE}
race.form <- WomenPerClinic ~ PercWhite + PercBlack + PercAmInd + PercAsian + PercHawaii + PercTwoOrMore + PercHisp

race.spatial.model <- S.CARleroux(formula=race.form, data=tn.spatial, family="gaussian", W=W, burnin=100000, n.sample=300000, thin=20)
```

```{r}
print(race.spatial.model)
```

Our intuitive takeaways from this model: 

 * White and American Indian are the only groups that cause a *decrease* in women per clinic -- I wonder if the American Indian phenomenon follows the same logic as the rural/high school grad coefficients in the previous model. I don't think there are that many counties with significant native populations in Tennessee, and I'd guess they are rural, undereducated counties. 
 * Asian and Hawaiin/Pacific Islander cause the greatest increase in women per clinic. 

I think that these coefficients will change if we fit the model to the entire country, but the signs and intuition will probably stay the same. 

## Progress since last week
This was my minimal progress week! I fit the models here, read the original papers about S.CARdissimilarity and S.CARleroux, and spent *a lot* of time unsuccessfully trying to fit models to the entire country. Oh well!

